---
title: "PulseWaveformTutorial2"
author: "Simon Williamson"
date: "04/07/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PulseWaveform 

This tutorial will use an example PPG time series to demonstrate the PulseWaveform pipeline for PPG feature extraction and running of the HED model (see github Readme for an introduction to the model and pipeline). The example data is a a two-minute segment of raw data outputted by a BioRadio device at 75 Hz. 

# Initial setup and loading data

### Install PulseWaveform

If not done so already, install the PulseWaveform package with:

```{r message=FALSE}
devtools::install_github(repo = 'stw32/PulseWaveform')
```

### Load required packages:

These packages may need to be installed first.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(splines2)
library(pracma)
library(SplinesUtils)
library(spectral)
library(DescTools)
library(zoo)
library(readr)
library(PulseWaveform)
```

### Select starting parameters

In this case we know the sampling rate of Bioradio data (75 hz), and that most primary (but not secondary) PPG peaks in the first derivative are above 0.8 on the y-axis (this is important for proper peak detection). We also wish to visualise the segmented waveforms, and how the extracted fiducial points 'OSND' vary across the sample (both plotting parameters are therefore set to true):

```{r}
samplingRate <- 75                
pk_thrshd <- 0.8                    
plot_aligned_waves <- TRUE      
plot_osnd <- TRUE    
```

Following this we select parameters for running the model. Let's say that we do want to run the model, with the recommended number of simplex iterations. We want all beats in the sample to be modelled (not just a subset), making the 'batch_number' parameter redundant. 

'beats_in' defines the number of beats to be optimised with some shared parameters (see Readme). It is effectively a way to alter the trade-off between consistency of model fits (and therefore robustness) vs goodness of fit. In this case we will set it to 2:

```{r}
run_hed <- TRUE                     
simplex_iterations <- 20000        
all_beats <- FALSE                   
batch_number <- 6                 
beats_in <- 2                    
```

### Download example data

Download the sample data to a temporary directory and then load into R (avoiding directory related issues...)):

```{r}
tmpDir = tempdir(check = TRUE)
download.file("https://github.com/stw32/PulseWaveform/blob/main/Example%20Data/example_PPG_time_series.csv?raw=true", 
              destfile = file.path(tmpDir, "sample_data.csv"))
data <- read.csv(file.path(tmpDir, "sample_data.csv"), header = T)
```

### Define output folder

Next we define the output folder for all extracted features and segmented waveforms the pipeline will generate:

```{r}
AllOutputs <- list()   
```


# Preprocessing

First duplicates are removed from the data, and the undetrending algorithm inherent to Bioradio hardware is reversed. The raw data is assigned to a dataframe 'ppg', and an initial attempt at detecting peaks in the first derivative is made (to be refined later). 

```{r}
undetrended_data <- data.frame(preproc(dat=data))                                                  


ppg <- undetrended_data                                                                          
ppg <- data.frame(
  time = (0:(nrow(ppg)-1)) / samplingRate,
  ppg = ppg[,1]
)
names(ppg)[1] <- "time (s)"
names(ppg)[2] <- "Detrended"


n <- dim(ppg)[1]                                                                                   
vpg <- ppg[2:n,2] - ppg[1:(n-1),2]                                                       
beat <- data.frame(ppg[which(vpg[1:(n-1)] < pk_thrshd & vpg[2:n] >= pk_thrshd),1])
rm(vpg)
undetrended <- ppg[, 2]
```

The preprocessed time series can now be visualised:

```{r}
plot((1:length(undetrended))/samplingRate, undetrended, t = "l", xlab = "Time (seconds)", ylab = "PPG (a.u.)")
```

# Extract Spectral Features

If desired, frequency domain features can be extracted from the PPG time series at this point. These include low, medium and high frequency band powers (relative to total power), as well as low to high frequency ratio. Frequency bands are defined according to Lee et al, 2011 [https://pubmed.ncbi.nlm.nih.gov/21693795/]. 



```{r}
N <- length(undetrended)
xPer <- (1/N)*abs(fft(undetrended)^2)
f <- seq(0,1.0-1/N,by=1/N)
f <- f*samplingRate
spectrum <- data.frame(f, xPer)
point04 <- which(abs(f - 0.04) == min(abs(f - 0.04)))
point08 <- which(abs(f - 0.08) == min(abs(f - 0.08)))
point145 <- which(abs(f - 0.145) == min(abs(f - 0.145)))
point45 <- which(abs(f - 0.45) == min(abs(f - 0.45)))
total_power <- sum(spectrum[point04:point45, 2])
LFNU <- sum(spectrum[point04:point145, 2]) / total_power
MFNU <- sum(spectrum[point08:point145, 2]) / total_power
HFNU <- sum(spectrum[point145:point45, 2]) / total_power
LFHF_ratio <- LFNU / HFNU
spectrum[c(which(1:nrow(spectrum) < point04)), ] <- 0
plot(spectrum, t = "l", xlim = c(0.04, 1.5), ylim = c(0, max(spectrum$xPer)), xlab = "frequency", ylab = "power")
print(LFHF_ratio)
```

# Beat Segmentation

In this section, peaks in the time series are formally identified and used as the basis for segmentation of the time series into individual waveforms. 

```{r results="hide"}
undetrended <- ppg[, 2]
sfunction <- splinefun(1:length(undetrended), undetrended, method = "natural")
deriv1 <- sfunction(seq(1, length(undetrended)), deriv = 1)
spline1 <-  sfunction(seq(1, length(undetrended)), deriv = 0)
splinePoly <- CubicInterpSplineAsPiecePoly(1:length(undetrended), undetrended, "natural")
deriv1Poly <- CubicInterpSplineAsPiecePoly(1:length(undetrended), deriv1, "natural")
inflexX <- solve(splinePoly, b = 0, deriv = 1)
inflexY <- predict(splinePoly, inflexX)
w <- find_w(d1p = deriv1Poly, deriv1 = deriv1, sp = splinePoly, sr = samplingRate, pk_thrshd = pk_thrshd) 
```

The identified peaks can then be visualised as found in the first derivative of the time series:

```{r}
plot(deriv1, t = "l", ylab = "PPG first derivative")
points(w$wX, w$wYD1, col = "red")
```

The time series is then segmented into individual waveforms. Cleaning steps occur pre and post segmentation. 

```{r results="hide"}
uv <- find_u_v(wx = w$wX, wy = w$wY, d1 = deriv1, d1p = deriv1Poly,
               spline = splinePoly, sr = samplingRate, plot=F)
tmp <- find_o(wx = w$wX, inx = inflexX, iny = inflexY, d1p = deriv1Poly, sp = splinePoly)       
inflexX <- tmp[[1]]
inflexY <- tmp[[2]]
o_orig <- tmp[[3]]
tmp <- preclean_wuv(w=w, uv=uv, o=o_orig, samp = samplingRate, sp = spline1, q = F)
w <- tmp[[1]]
uv <- tmp[[2]]
o <- tmp[[3]]
rm(tmp)
baseCor <- baseline(inx = inflexX, iny = inflexY, o = o_orig,                                 
                    dat = undetrended, sp = splinePoly, plot=F)
sfunctionBC <- splinefun(1:length(baseCor), baseCor, method = "natural")
deriv1BC <- sfunctionBC(seq(1, length(baseCor)), deriv = 1)
spline1BC <- sfunctionBC(seq(1, length(baseCor)), deriv = 0)
splinePolyBC <- CubicInterpSplineAsPiecePoly(1:length(baseCor), baseCor, "natural")
deriv1PolyBC <- CubicInterpSplineAsPiecePoly(1:length(baseCor), deriv1BC, "natural")
w$wY <- predict(splinePolyBC, w$wX)
uv$uY <- predict(splinePolyBC, uv$uX)
uv$vY <- predict(splinePolyBC, uv$vX)
wuv <- cbind(w, uv)
tmp <- clean_wuv(wuv = wuv, sp = splinePolyBC, inx = inflexX, o = o,
                 samp = samplingRate, bc = baseCor, q = F)
wuv <- tmp[[1]]
ibi <- tmp[[2]]                                                                                   
oDiff <- tmp[[3]]
rm(tmp, w, uv)
waveLen <- round(median(oDiff)+15)
ppg[, 2] <- baseCor
tmp <- sep_beats(odiff = oDiff, bc = baseCor, samp = samplingRate, wuv = wuv, wvlen = waveLen,     
                 ibi=ibi, o=o_orig, inx = inflexX, scale = T, q = F, subset = FALSE, boundaries)
avWave <- tmp[[1]]
pulse <- tmp[[2]]
wuv <- tmp[[3]]
rejects <- tmp[[4]]                                                                             
rm(tmp)
```

All pulse waveforms can then be aligned and visualised, with the average (mean) waveform overlayed in red:

```{r}
if(plot_aligned_waves == TRUE){                                                            
  pulse_stacked <- gather(pulse, key = "wave_ID", value = "values", -c("x"))
  average <- data.frame(seq((-141/(samplingRate*10)),
                            ((waveLen*15 -9)-142)/(samplingRate*10),
                            by = 1/(samplingRate*10)))
  average <- cbind(average, avWave)
  colnames(average)[1] <- "x"
  pl <- ggplot(data = pulse_stacked[-which(is.na(pulse_stacked[, 3])), ],
               aes(x, values, col = wave_ID), col = "black") +
    scale_color_manual(values = rep("black", ncol(pulse))) +
    geom_line(size = 1.5, alpha = ((1/length(wuv$wX)*10)-(1/length(wuv$wX)))) +
    geom_line(data = average[-which(is.na(average[, 2])), ],
              aes(x, avWave), size = 1.125, color = "red") +                                     
    theme(legend.position = "none") + labs( y= "PPG Output", x = "Time (Seconds)") +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"),
          axis.title = element_blank(),
          axis.text = element_text(size = 12)) 
  print(pl)
}
```

# The HED Model

Prior to running the HED model, starting parameters must be estimated. The starting parameter for 'decay rate' (i.e. config.rate) is defined first. 0.99 is recommended for canonical waveforms, lower values may be required for non-canonical waveforms e.g class 3 or above (for description of classes, see figure 1 of Tigges et al, 2016: https://www.degruyter.com/document/doi/10.1515/cdbme-2016-0046/html?lang=en). 

```{r}
config.rate = 0.99  
lab.time = "time (s)"
lab.ppg = "Detrended"
const.pi = 3.1415926535897932384626433
```

As the HED model runs it will generate plots of fitted waveforms. It is worth viewing these to assess the overall quality and plausability of fits. For the sake of space, however, we will only plot two here by ensuring the 'PlotFits' function only runs once: 

```{r}
plotfitsgo = TRUE
```


Running HED (this is the most computationally demanding section of the pipeline and may therefore take some time). 

```{r results="hide"}
if(run_hed == TRUE){
  
  beat <-  ppg[round(inflexX[wuv$o2]), 1]                                                          
  nBeats <- length(beat)
  beat <- data.frame(
    beat = beat,
    dt = (1:nBeats)*0.0
  )
  beat <- AddOutput(beat)
  if(all_beats == T){
    batch_number <- floor(nrow(beat)/beats_in)
    remainder <- nrow(beat) - (batch_number*beats_in)
  }
  
  ppg$Baseline = 1:nrow(ppg) * 0                                                                   
  ppg$Excess   = 1:nrow(ppg) * 0
  ppg$Residue  = 1:nrow(ppg) * 0
  
  temp <- FindStartParams(batch_number, beats_in, beat, ppg, gs = model2.GetSegment,               
                          e = model2.Excess, sep = model2.SubtractExcessPeak,
                          o_points = inflexX[o_orig], wuv = wuv, inflexX = inflexX, all_beats)
  beat <- temp[[1]]
  ppg <- temp[[2]]                                                                                 
  rm(temp)
  
  beat_orig <- beat
  fit_check <- list()
  for(k in 1:(batch_number+1)){
    
    if(all_beats == TRUE){
      if(k == batch_number+1){
        if(remainder == 0){break}
        beat <- beat_orig[(((k-1)*beats_in) + 1 ):(((k-1)*beats_in) + remainder), ]
        beats_in <- remainder
        w <- wuv$wX[(((k-1)*beats_in) + 1 ):(((k-1)*beats_in) + remainder)]
      }else{
        beat <- beat_orig[((k*beats_in)-(beats_in-1)):(k*beats_in), ]
        w <- wuv$wX[((k*beats_in)-(beats_in-1)):(k*beats_in)]
      }
    }else{
      if(k == batch_number+1){break}
      beat <- beat_orig[((k*beats_in)-(beats_in-1)):(k*beats_in), ]
      w <- wuv$wX[((k*beats_in)-(beats_in-1)):(k*beats_in)]
    }
    
    w <- w / samplingRate
    renal_param <- median(beat$NTime)
    dias_param <- median(beat$DTime)
    sys_time <- beat$STime
    par <- as.numeric(beat[1,5:16])
    beat_start <- beat[, 3]
    beat_end <- beat[, 4]
    beat_vector <- list(beats_in, beat_start, beat_end)
    
    for(i in 1:4){                                                                              
      if(i == 1){new_beat <- beat}
      within_params <- FindWithinParams(beats_in, ppg, beat = new_beat, gs = model2.GetSegment,
                                        fp = model2.FixParams3, ms = simplex.MakeSimplex3,
                                        m2 = model2.ChiSq3, beat_vector = beat_vector,
                                        renal_param = renal_param, dias_param = dias_param,
                                        sys_time = sys_time, w = w)
      across_params <- simplex.MakeSimplex2(data=ppg, param = par, f = model2.ChiSq3,
                                            inScale = 0.1, beat_vector = beat_vector,
                                            beat = new_beat, renal_param = renal_param,
                                            dias_param = dias_param, sys_time = sys_time, w = w)
      mat <- make_matrix(across_params, within_params)
      sim <- simplex.Run2(data = ppg, simplexParam = mat, f = model2.ChiSq3, optional=NULL,
                          beat_vector = beat_vector, renal_param = renal_param,
                          dias_param = dias_param, sys_time = sys_time, ms = simplex_iterations,
                          w = w, run = c("run", i))
      output <- extractOutput(beats_in, sim)
      fixed <- FixOutput(beats_in, beat = new_beat, ppg, gs = model2.GetSegment,
                         fp = model2.FixParams3, across = output[[1]], within = output[[2]],
                         sys_time = sys_time)
      new_beat <- UpdateBeat(beats_in, beat, fixed)
      new_beat <- FixBaseline(new_beat, f = model2.ChiSq4, renal_param, dias_param, sys_time, w)
    }
    
    
    fit_check[[k]] <- model2.ChiSq4(data = ppg, params = NULL, beats = beat_vector,                
                                    beat = new_beat, a = sim[1, ], plot = FALSE,
                                    renal_param = renal_param, dias_param = dias_param,
                                    sys_time = sys_time, w = w)
    
    beat2 <- new_beat                                                                              
    colnames(beat2) <- colnames(beat)
    beat2 <- beat2[, -c(1:4)]
    
    
    if (plotfitsgo == T){
      PlotFits(beats_in, ppg, beat2, gs = model2.GetSegment, rb = model2.Rebuild2)                 
    }
    
     if (plotfitsgo == T){
     GGplotFits(beats_in, ppg, beat2, gs = model2.GetSegment, rb = model2.Rebuild2,         
     run = 1, pr = 1, p = T, iso = F)
     #plotfitsgo <- FALSE
     }
    
    if(k == 1){beat_final <- beat2}else{beat_final <- rbind(beat_final, beat2)}                    
    
  }
}
```


# Feature Extraction

### Morphological feature extraction

First fiducial points (OSND) are identified on each individual waveform. 

```{r results="hide"}
polyWave <- list()                                                                                 
for(i in 2:ncol(pulse)){
  polyWave[[i-1]] <-CubicInterpSplineAsPiecePoly(pulse$x, pulse[, i], "natural")
}
tmp <- diast_pk(avw = avWave, sr = samplingRate, scale = T)
dPeak <- tmp[1]
xShift <- tmp[2]
rm(tmp)
osnd <- osnd_of_average(avWave, dp = dPeak, diff = 0, sr = samplingRate, plot = F)
if(dPeak == 5*samplingRate){
  dPeak <- osnd$x[4]*1.2
}
if((osnd$x[4]-osnd$x[3]) < 1.5 & (osnd$x[4]-osnd$x[3]) > 0){
  dPeak <- dPeak*0.95
  osnd <- osnd_of_average(avWave, dp = dPeak, diff = 0, sr = samplingRate, plot = FALSE)
}
scale <- 1                                                                                         
osnd_all <- list()
for(i in 2:ncol(pulse)){
  wavi <- pulse[, i][!is.na(pulse[, i])]
  if(scale == 1){
    xShift2 <- (which(abs(wavi - 0.5) == min(abs(wavi - 0.5))))
  }else{
    xShift2 <- which.min(abs(wavi))
  }
  diff <- xShift - xShift2
  dpa <- dPeak - diff
  osnd_all[[i-1]] <- osnd_of_average(aw = wavi, dp = dpa, diff = diff,
                                     sr = samplingRate, plot = F)
}
```


These can then be plotted against the average waveform:

```{r, results = "hide"}
if(plot_osnd == TRUE){                                                                             
  plot(avWave[!is.na(avWave)], type = "l") + for(i in 1:length(osnd_all)){
    points(osnd_all[[i]][4, 1], osnd_all[[i]][4, 2], col = "blue")
    points(osnd_all[[i]][3, 1], osnd_all[[i]][3, 2], col = "red")
    points(osnd_all[[i]][2, 1], osnd_all[[i]][2, 2])
    points(osnd_all[[i]][1, 1], osnd_all[[i]][1, 2])
  }
}
```

Then features derived from fiducial points (e.g. augmentation index) can be derived:

```{r results="hide"}
for(i in 1:length(osnd_all)){                                                                      
  osnd_all[[i]]$y <- osnd_all[[i]]$y - osnd_all[[i]]$y[1]
}
features <- feature_extract(oa = osnd_all, p = pulse, pw = polyWave)
```

### Model Outputs finalised

```{r results="hide"}
if(run_hed == TRUE){
  beat_final <- cbind(beat_orig[1:nrow(beat_final), 1:4], beat_final)                              
  osnd_fits <- osnd_fit(beat_final, ppg, plot = F)                                                 
}
```

# Outputs

Finally, the outputs generated across the pipeline are arranged into the 'AllOutputs' list:

```{r}
if(run_hed == FALSE){
  fit_check <- list(c(1:100), c(1:100), c(1:100))
  beat_final <- data.frame(1:nrow(features))
  beat_orig <- data.frame(1:nrow(features))
  osnd_fits <- list(1:100)
  }
temp <- ArrangeOutputs(beat_final, beat_orig, features, pulse, fit_check)
nBeats <- ncol(pulse) - 1
AllOutputs <- list(nBeats, ibi, rejects, pulse, polyWave, osnd_all, avWave, osnd, temp[[2]], temp[[1]], temp[[3]], osnd_fits)
```

The key to the output list is as follows:

```{r}
# [[1]] == nBeats                         Number of waveforms in sample
# [[2]] == ibi                            Inter-beat intervals (note pre-interpolation)
# [[3]] == rejects                        Rejected beats, arranged according to reason for exclusion (see Readme)
# [[4]] == pulse                          Individual waveforms in the sample (discrete form)
# [[5]] == polyWave                       Individual waveforms in the sample (polynomial spline form)
# [[6]] == osnd_all                       OSND points of individual waves
# [[7]] == avWave                         Average waveform of the sample
# [[8]] == osnd (of average)              OSND points of the average waveform
# [[9]] == features                       Morphological features derived from OSND points (see supplementary material)
# [[10]] == beat_final                    Model parameter Outputs
# [[11]] == fit_check                     Goodness of fit Measures (ChiSq, Max error, NRMSE, aNRMSE)
# [[12]] == osnd_fits                     Error values (x and y) in recapitulation of OSND points
```


# Exploring Outputs

Once the pipeline has run, outputs can be examined in a number of ways. Below of some examples of visualisations that may be of interest. Outputs that prove to be informative could also be used as the basis for classification models. 

### Interbeat intervals

An interbeat interval time series is shown with the following plot. Note that due to cleaning and removal of waveforms, caution should be taken when interpreting: intervals are not necessarily. Nonetheless, an approximation to heart rate variability across the time series can be extracted by taking the standard deviation of interbeat intervals (akin to SDNN) after removal of outliers. 


```{r}
plot(AllOutputs[[2]], t = "l")
```

In this plot we see a single outlier value, likely due to an interval between non-consecutive peaks. It should be removed before calculating heart (pulse) rate variability:

```{r}
ibis <- which(AllOutputs[[2]] < 80)
sdrr <- sd(ibis)
print(sdrr)
```


### Morphological features

Notch height (relative to systolic peak height) is a feature that is likely associated with sympathetic arousal.

### HED Model Performance

#### Goodness of fit

We can assess goodness of fit across the sample of waveforms by visualising NRMSE values for each fitted waveform. In this case the median NRMSE is 0.9 (2dp). 

```{r}
hist(AllOutputs[[11]][[3]], main = "Histogram of NRMSE values", xlab = "NRMSE")
print(median(AllOutputs[[11]][[3]]))
```


### Exploring HED model outputs

HED Model outputs can also be plotted as time series. This many help to elucidate, for instance, 

#### Observation of dynamic beat-to-beat waveform behaviour

This animation may be useful for visualising how waveform morphology (and subsequent modeling) changes on a beat to beat basis. 

# End

This is the end of the tutorial. If there are any questions please do refer to the Readme. To apply the pipeline to your own PPG data, the *general purpose script containing the code above is available and can be altered as required for multiple time series / individuals / groups. 



